%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%%  Curriculum Vitae
%%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

\markboth{Curriculum Vitae}{}
\chapter*{Curriculum Vitae}
\addcontentsline{toc}{chapter}{Curriculum Vitae}


\begin{tabularx}{\textwidth}{>{\raggedleft}p{0.285\textwidth}X}
	Name  & \thesisauthor \\
	& \thesisauthoraddress \\
	Date of Birth & \thesisauthordateofbirth \\
	Birthplace & \thesisauthorbirthplace \\
	Citizenship & \thesisauthornationalityCv \\
\end{tabularx}

\section*{Education}
\begin{tabularx}{\textwidth}{>{\raggedleft}p{0.285\textwidth}X}
	since Oct. 2017  & Ph.\,D. in Computer Science under the supervision of Prof.\,Dr.\,Heiko Schuldt, Databases and Information Systems Tesearch Group, University of Basel, Switzerland \\
	\rule{0pt}{3ex}%  EXTRA vertical height
	Sep. 2014 -- Jul. 2017  & M.\,Sc. in Computer Science, University of Basel, Switzerland \\
	\rule{0pt}{3ex}%  EXTRA vertical height
	Sep. 2007 -- Jul. 2014  & B.\,Sc. in Nanoscience, University of Basel, Switzerland \\
	\rule{0pt}{3ex}%  EXTRA vertical height
	Sep. 2002 -- Dec. 2006  & Matura, Gymnasium Liestal, Liestal, Switzerland \\
\end{tabularx}

\section*{Employment}
\begin{tabularx}{\textwidth}{>{\raggedleft}p{0.285\textwidth}X}
	since Oct. 2017  & Research \& Teaching Assistant, Databases and Information Systems Research Group, University of Basel, Switzerland \\
	since Dec. 2010  & IT Consultant \& Partner, pontius software GmbH, Bubendorf, Switzerland \\
	Jun. 2008 - Dec. 2011  & Application Manager ``imdas pro'', Bildungs-, Kultur- und Sportdirektion des Kantons Basel-Landschaft, Liestal, Switzerland
\end{tabularx}

\newpage

\section*{Publications}

\paragraph{2022}
\begin{itemize}
	\item \fullcite{Heller:2022Multi}
	\item \fullcite{Spiess:2022Multi}
\end{itemize}

\paragraph{2021}
\begin{itemize}
	\item \fullcite{Rossetto:2021A}
	\item \fullcite{Heller2021:Interactive}
	\item \fullcite{Spiess2021:Exploring}
	\item \fullcite{Heller:2021Towards}
	\item \fullcite{Spiess2021:Competitive} 
	\item \fullcite{Rossetto:2021On}
\end{itemize}

\paragraph{2020}
\begin{itemize}
	\item \fullcite{Gasser:2020Cottontail} (Best Open Source System Award)
	\item \fullcite{Boerlin:20203d}
	\item \fullcite{Heller2020:Interactive}
	\item \fullcite{Rossetto:2020Interactive}
	\item \fullcite{Sauter2020:Combining}
\end{itemize}

\paragraph{2019}
\begin{itemize}
	\item \fullcite{Gasser:2019Multimodal} (Best Demo Award)
	\item \fullcite{ Giangreco2019:Virtue}
	\item \fullcite{Rossetto2019:Retrieval}
	\item \fullcite{Lokovc:2019Interactive}
	\item \fullcite{Rossetto:2019Deep}
\end{itemize}
\paragraph{2018}
\begin{itemize}
	\item \fullcite{Rossetto:2018Competitive}
\end{itemize}

\section*{Advised Theses}

\begin{itemize}
	\item Colin Saladin. \emph{SIMD Support for Proximity Based Queries in Cottontail DB}. Bachelor's Thesis, University of Basel
	\item Renato Farrugio. \emph{Scene Text Recognition in Images and Video with Cineast}. Bachelor's Thesis, University of Basel
	\item Florian Mohler. \emph{Online Image Analysis to Identify Politicians in Twitter Images}. Bachelor's Thesis, University of Basel
	\item Gabriel Zihlmann. \emph{Magnetic Resonance Fingerprinting Reconstruction using Methods from Multimedia Retrieval}. Master's Thesis, University of Basel
	\item Yan Wang. \emph{Detecting Visual Motives Using Online Clustering of Similar Images}. Bachelor's Thesis, University of Basel
	\item Samuel Börlin. \emph{3D Model Retrieval using Constructive Solid Geometry in Virtual Reality}. Master's Thesis, University of Basel
	\item Manuel Hürbin. \emph{Retrieval Optimization in Magnetic Resonance Fingerprinting}. Bachelor's Thesis, University of Basel
	\item Loris Sauter. \emph{Fighting Misinformation: Image Forgery Detection in Social Media Streams}. Master's Thesis, University of Basel
\end{itemize}

\section*{Honors and Awards}

\begin{description}
	\item [Best VBS System Award] for the winning system of the \acrshort{vbs} and paper titled \emph{Towards Explainable Interactive Multi-Modal Video Retrieval with vitrivr} at the \emph{27th International Conference on Multimedia Modeling (MMM 2021)}
	\item [Best Demo Award] for the best demo paper titled \emph{A System for Interactive Multimedia Retrieval Evaluations} at the \emph{27th International Conference on Multimedia Modelling (MMM 2021)}
	\item [Best Open Source Award] for the best open source system paper titled \emph{Cottontail DB: An Open Source Database System for Multimedia Retrieval and Analysis} at the \emph{28th ACM International Conference on Multimedia (ACM MM 2020)}
	\item [Best Demo Award] for the best demo paper titled \emph{Multimodal Multimedia Retrieval with vitrivr} at the \emph{ACM International Conference on Multimedia Retrieval (ICMR 2019)}
	\item [Best VBS System Award] for the winning system of the \acrshort{vbs} and paper titled \emph{Deep Learning-based Concept Detection in vitrivr} at the \emph{25th International Conference on Multimedia Modeling (MMM 2019)}
	\item [Fritz Kutter Award 2017] for the Master's Thesis titled \emph{Towards an All-Purpose, Content-based Multimedia Retrieval System}. The Fritz Kutter trust (managed by ETH Zürich) honours the best thesis at a Swiss University every year.
\end{description}

\section*{Other Relevant Experience}

\begin{itemize}
	\item Technical Program Committee (PC) Member of the \emph{International Workshop on Interactive Multimedia Retrieval (IMuR	
	2022)} 
	\item Technical Program Committee (PC) Member of the \emph{International Conference on Multimedia Modeling (MMM 2022)} 
	\item Technical Program Committee (PC) Member of the \emph{Fourth Annual Workshop on Lifelog Search Challenge (LSC 2021)} 
	\item Technical Program Committee (PC) Member of the \emph{Second International Workshop on Video Retrieval Methods and Their Limits (ViRaL 2021)}
	\item Technical Program Committee (PC) Member of the \emph{ACM International Conference on Multimedia (ACM MM 2020)} 
\end{itemize}

\cleardoublepage
