%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%%  ACKNOWLEDGEMENTS
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

\chapter{Acknowledgements}

Finishing a project such as the one described in this Thesis is never the sole achievement of one person alone. And while I may be the author of the text you read and the mind behind the concepts and systems you read about, there is a great many people that have somehow influenced the completion of this dissertation. It is those sung and unsung heroes that I shall dedicate these acknowledgements to.

First and foremost, I would like to thank Prof. Dr. Heiko Schuldt, who offered me this position and who somehow convinced me five years ago, that doing a PhD is a unique opportunity better not to be missed. The past five years indeed were time very well spent on interesting projects with interesting people and I am grateful for the experience and the guidance you have given me along this journey. I also wish to thank Assc. Prof. Dr. Björn Þór Jónsson, who volounteered as my external examinor. Reading this Thesis of mine must have been a daunting task indeed and I very much hope that the experience was at least somewhat pleasant.

I would also like to take the opportunity to give a big shoutout to our DBIS research group with all the amazing members both past and present. I really enjoyed our many lunches, discussions about politics, progress reports and of course the mandatory gossip on the second floor. A special thanks goes to Silvan Heller, Loris Sauter, Marco Vogt, Alexander Stiemer, Florian Spiess and Mahnaz Parian-Sherb as well as our former members Luca Rossetto and Ivan Giangreco who I do not only consider valued colleagues but also dear friends.  

The process of writing and finishing-up a Thesis like this can be taxing and downright frustrating at times. In this situation, it is valuable to have colleagues that share the same predicament and with whom one can exchange about the progress, the set backs and just the general state of mind. Therefore a loud, emphatic ``thank you'' to both Marco and Silvan for establishing and holding onto our ``Thesis Reading Circle'' and for the many hours of discussion and the invaluable feedback you have provided.

Team DRES (Loris \& Luca) also deserves a special mention: It is not often that one finds people with whom to work is both as productive and pleasant as with the two of you and really would like to thank you for this. We have achieved a lot together and I am positive that there are still many interesting side-projects to be started, papers to be written and discussions to be had.

I would also like to acknowledge and thank my dear friend Andrea Schäfer for reading my entire Thesis and providing me with feedback on the English language. I guess without you, reading this damn thing would be even more painful than it probably is anyway. And I also acknowledge Luca Rossetto for providing feedback on bits and pieces every once in a while.

Of course, I also want to thank my friends and family -- all those of you who allowed and enabled me to wind down from the at times stressful experience of working on this project. A heartfelt ``thank you'' and, once again, a big shoutout to all of you -- also for the patience you've shown over the past few years.

Now there is one person that deserves a special mention here -- my dear and beloved companion, Jonas. We have had over thirteen years together but the past five years probably were among the most challenging. Balancing two occupations, the buying and rennovation of a house, the well-being of a herd of rabbits and life in general certainly isn't an easy task and it would not have been possible without your unwavering support. Thank you for being there for me -- in good and in bad times, as they say -- and for strengthening my back regardless of which adventure I embark on.

And last but not least, thanks a lot to everyone who has not been mentioned explicitly yet is explicitly meant: The students and collegues that I had the honour of working with, the administrative staff that makes sure that everything runs smoothly and all the generation of researchers who's efforts have enabled this small contribution that I'm trying to make with this Thesis. Thanks to everyone, who has supported me directly or indirectly, knowling or unknowingly, willingly or unwillingly -- thank you very much!

\epigraph{We learned about gratitude and humility - that so many people had a hand in our success, from the teachers who inspired us to the janitors who kept our school clean and we were taught to value everyone's contribution and treat everyone with respect.}{--- \textup{Michelle Obama}, 2012}


\cleardoublepage
